Hockey Enforcers

### 2024-02-15 HocketEnforcersReport.Rmd
Updated.

This is the exact number of players that the paper cites.  
```{r}
#This is the number they get in the paper.  
gen_JAMA %>% filter(X1st.Season.Corrected >= 1967) %>% nrow()
```

@@

#All these goalies.
dead$Name[!dead$Name %in% gen_JAMA$Player]
```

```{r}
#Dataprep
dat <- dat %>% mutate(dead = !is.na(Death.Date), 
                      DOB = as.Date(DOB), 
                      Death.Date = as.Date(Death.Date))
#I'm using the date of the latest death date as the censoring time: May 7, 2023
dat$Death.Date[!dat$dead] <- as.Date("2023-05-07")
dat$age <- (dat$Death.Date - dat$DOB)/365
dat$Ht <- as.numeric(dat$Ht)
dat$Wt <- as.numeric(dat$Wt)
dat$FperG <- dat$Total.Fights/dat$GP.x
dat$Pos <- as.factor(dat$Pos)
dat$Ctry2 <- dat$Ctry
dat$Ctry2[!dat$Ctry2 %in% c("FIN","CZE","RUS","SWE","USA","CAN")] <- "OTH" 
dat$Ctry2 <- as.factor(as.character(dat$Ctry2))
dat$EF <- dat$Total.Fights>50

dat <- dat %>% filter(!is.na(Ht))
```

Creating the Surv object
```{r}
library(survival)
#Defining the survival variable
dat$Surv <- Surv(dat$age,dat$dead)
```
Some plots.
```{r}
ggplot(aes(x = age,col = EF), data = dat) + geom_density()
ggplot(aes(x = Wt,col = EF), data = dat) + geom_density()
ggplot(aes(x = Wt,y = age,col = EF), data = dat) + geom_point() + geom_smooth()
```
Some more plots.
```{r}
ggplot(aes(x = Total.Fights,y = age), data = dat) + geom_point() + geom_smooth()
ggplot(aes(x = Total.Fights,y = age, col = dead), data = dat) + geom_point() + geom_smooth()
ggplot(aes(y = Total.Fights,x = Wt), data = dat) + geom_point() + geom_smooth()
```

@@

# Cox model with Total.fights as a covariate.  
```{r}
a <- coxph(Surv ~ Total.Fights + GP.x + Ht + Wt + DOB.Year + Pos , data = dat)
summary(a)
```
# Cox model with EF as a covariate.  
No significant effect for enforcers in terms of survival times.  
Though the sign of the effect is what you would expect to see if there was a risk to being an enforcer.

```{r}
a <- coxph(Surv ~ EF + GP.x + Ht + Wt + DOB.Year + Pos , data = dat)
summary(a)
```

@@

library(survival)
#Create a survival variable on the matched data.  
dataMatched$Surv <- Surv(dataMatched$age,dataMatched$dead)
```

# Effect of Enforcer on hazard ratio after matching
n = 640 here.  
```{r}
coxph(Surv ~ EF, data = dataMatched)
```

```{r}
coxph(Surv ~ EF + GP.x + Ht + Wt + DOB.Year + Pos, data = dataMatched)
```
# Conditioning on dead people

```{r}
dataMatched %>% filter(dead) %>% group_by(EF) %>% summarise(mnage = mean(as.numeric(age)), mnwt = mean(Wt), mnDOB = mean(DOB.Year), mngp = mean(GP.x))

>>>>

This is the exact number of players that the paper cites.  
```{r}
#This is the number they get in the paper.  
gen_JAMA %>% filter(X1st.Season.Corrected >= 1967) %>% nrow()
#I get 331 individuals with greater than 50 fights
gen_JAMA %>% filter(X1st.Season.Corrected >= 1967 & Total.Fights >= 50) %>% nrow()
```

@@

#Remove goalies 
#There is no one to remove because there are no goalies in the original data set.  
dat <- dat %>% filter(Pos != "G")

#All these goalies.
dead[!dead$Name %in% gen_JAMA$Player,]
```

```{r}
#Dataprep
dat <- dat %>% mutate(dead = !is.na(Death.Date), 
                      DOB = as.Date(DOB), 
                      Death.Date = as.Date(Death.Date))
#I'm using the date of the latest death date as the censoring time: May 7, 2023
dat$Death.Date[!dat$dead] <- as.Date("2023-05-07")
dat$age <- (dat$Death.Date - dat$DOB)/365
dat$Ht <- as.numeric(dat$Ht)
dat$Wt <- as.numeric(dat$Wt)
dat$FperG <- dat$Total.Fights/dat$GP.x
dat$Pos <- as.factor(dat$Pos)
dat$Ctry2 <- dat$Ctry
dat$Ctry2[!dat$Ctry2 %in% c("FIN","CZE","RUS","SWE","USA","CAN")] <- "OTH" 
dat$Ctry2 <- as.factor(as.character(dat$Ctry2))
#At least 50, not greater than 50.
dat$EF <- dat$Total.Fights>=50

#Cleaning penalty minutes
pm_temp <- data.frame(do.call(rbind,strsplit(dat$PIM.GP,":")))
dat$PIM.GP_numeric <- as.numeric(pm_temp[,1]) + as.numeric(pm_temp[,2])/60
dat$EP <- dat$PIM.GP_numeric >=3
#I match the paper's n = 183 EP.But when I remove the people with missing height, I lose one of the 182 and end up with 182.

#Removed 4 rows
dat <- dat %>% filter(!is.na(Ht))

Creating the Surv object
```{r}
library(survival)
#Defining the survival variable
dat$Surv <- Surv(dat$age,dat$dead)
```
Some plots.
```{r}
ggplot(aes(x = age,col = EF), data = dat) + geom_density()
ggplot(aes(x = Wt,col = EF), data = dat) + geom_density()
ggplot(aes(x = Wt,y = age,col = EF), data = dat) + geom_point() + geom_smooth()
```

Some plots.
```{r}
ggplot(aes(x = age,col = EP), data = dat) + geom_density()
ggplot(aes(x = Wt,col = EP), data = dat) + geom_density()
ggplot(aes(x = Wt,y = age,col = EP), data = dat) + geom_point() + geom_smooth()
```
Some more plots.
```{r}
ggplot(aes(x = Total.Fights,y = age), data = dat) + geom_point() + geom_smooth()
ggplot(aes(x = Total.Fights,y = age, col = dead), data = dat) + geom_point() + geom_smooth()
ggplot(aes(y = Total.Fights,x = Wt), data = dat) + geom_point() + geom_smooth()
```

@@

Enforcers - Penalty (EP) (blue), as defined by the paper, vs non-enforcers (red)
```{r}
library(ggsurvfit)
survfit2(Surv ~ EP, data = dat) %>% 
  ggsurvfit() +
  labs(
    x = "Year",
    y = "Overall survival probability"
  ) + 
  add_confidence_interval()
```

# Cox model with Total.fights as a covariate.  
```{r}
a <- coxph(Surv ~ Total.Fights + GP.x + Ht + Wt + DOB.Year + Pos , data = dat)
summary(a)
library(xtable)
xtable(summary(a)$coef)
```
# Cox model with EF as a covariate.  
No significant effect for enforcers in terms of survival times.  
Though the sign of the effect is what you would expect to see if there was a risk to being an enforcer.

```{r}
a <- coxph(Surv ~ EF + GP.x + Ht + Wt + DOB.Year + Pos , data = dat)
summary(a)
library(xtable)
xtable(summary(a)$coef)
```

```{r}
cor(dat[,c("PIM.GP_numeric","Ht","Wt","DOB.Year")])

a <- coxph(Surv ~ EP + GP.x + Ht + Wt + DOB.Year + Pos , data = dat)
summary(a)
library(xtable)
xtable(summary(a)$coef)

a <- coxph(Surv ~ PIM.GP_numeric  + GP.x + Ht + Wt + DOB.Year + Pos , data = dat)
summary(a)
library(xtable)
xtable(summary(a)$coef)
```

@@

library(survival)
#Create a survival variable on the matched data.  
dataMatched$Surv <- Surv(dataMatched$age,dataMatched$dead)

# Effect of Enforcer on hazard ratio after matching
#n = 640 here.

coxph(Surv ~ EF, data = dataMatched)
coxph(Surv ~ EF + GP.x + Ht + Wt + DOB.Year + Pos, data = dataMatched)

#Survival rates
chisq.test(dataMatched$EF, dataMatched$dead)
mean(dataMatched$dead[dataMatched$EF])
mean(dataMatched$dead[!dataMatched$EF])

library(ggsurvfit)
survfit2(Surv ~ EF, data = dataMatched) %>% 
  ggsurvfit() +
  labs(
    x = "Year",
    y = "Overall survival probability"
  ) + 
  add_confidence_interval()

Now repeat this with matching.  
```{r}
library(MatchIt)
#Mathing on EF
m.out <- matchit(EP ~ GP.x + Ht + Wt + DOB.Year + Pos, data = dat, ratio = 1)
dataMatched <- match.data(m.out)

library(survival)
#Create a survival variable on the matched data.  
dataMatched$Surv <- Surv(dataMatched$age,dataMatched$dead)

coxph(Surv ~ EP, data = dataMatched)
coxph(Surv ~ EP + GP.x + Ht + Wt + DOB.Year + Pos, data = dataMatched)

#Survival rates
chisq.test(dataMatched$EP, dataMatched$dead)
mean(dataMatched$dead[dataMatched$EP])
mean(dataMatched$dead[!dataMatched$EP])

library(ggsurvfit)
survfit2(Surv ~ EP, data = dataMatched) %>% 
  ggsurvfit() +
  labs(
    x = "Year",
    y = "Overall survival probability"
  ) + 
  add_confidence_interval()
```

# Conditioning on dead people

```{r}
dataMatched %>% filter(dead) %>% group_by(EF) %>% summarise(mnage = mean(as.numeric(age)), mnwt = mean(Wt), mnDOB = mean(DOB.Year), mngp = mean(GP.x), n = n())

dataMatched %>% filter(!dead) %>% group_by(EF) %>% summarise(mnage = mean(as.numeric(age)), mnwt = mean(Wt), mnDOB = mean(DOB.Year), mngp = mean(GP.x), n = n())

dataMatched %>% group_by(EF) %>% summarise(mnage = mean(as.numeric(age)), mnwt = mean(Wt), mnDOB = mean(DOB.Year), mngp = mean(GP.x), n = n())
